---
title: "timeseriesTrajectories"
author: "Thomas Mumuni Bilintoh"
date: "`r Sys.Date()`"
output: "github_document"
---

```{r setup,include = FALSE,warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width=8.5, 
  fig.height=6,
  comment = "#>"
)
# Make sure you load all the libraries below before you start using the package.
library(timeseriesTrajectories)
library(prettymapr)
library(progress)
library(RColorBrewer)
library(classInt)
library(ggnewscale)
library(reshape2)
library(terra)
library(dplyr)
library( data.table)
library(tmap)
library(tmaptools)
```

## Data
The example data is a time series of five binary maps at five time points: 2000, 2001, 2002, 2003, and 2005.
The tutorial analyzes the trajectory of variable 1, which in this instance indicates presence, whiles 0 represents absence.

The time series maps are available in GeoTIFF formats and included in the package. 
The first step is to load the raster files. 
```{r, eval=TRUE, warning=FALSE}
# Read the raster files from "externa"l folder as terra raster format.
#NB: The package comes with "Example_Data_Y.tif" and "Continuous_Data.tif".
#This tutorial shows results for only Example_Data_Y.tif.

rasstackY <- terra::rast(system.file("external/binary_raster_stack.tif",package="timeseriesTrajectories"))
rasstackY

# Visualize the raster data
plot(rasstackY)
```

## Calling the `dataClean` function
Next, use the **dataClean** function to 
reclassify your data if necessary. The functions build upon terra's **classify** function; thus, see terra's **classify** 
function for more information about creating the reclassification range. I will reclassify the raster stack, so that all 0s become 1s and 1s become 2s. 
to a binary raster stack to illustrate the functionality of the **dataClean** function.Type "?dataClean" in your R console to see the help file.

```{r, eval=TRUE, warning=FALSE}
# create a data frame for your range of values
reclass_df <- cbind(from = c(1), 
                    to = c(2), 
                    becomes = c(2))

#Use terra's **classify** function to reclassify the raster stacked based on your reclassification data frame.
# Setting "include.lowest = TRUE" ensures that the lower limit of the range is included in the values to be reclassified.
#  see terra's **classify** function for more information.
rasstackY_v2 <- classify(rasstackY,reclass_df,include.lowest = TRUE)

# Visualize the reclassified raster data
plot(rasstackY_v2)
```

## Calling the `dataPreview` function
The **dataPreview** function generates a time series graph for the variable where the vertical axis is the size of the variable, and the horizontal axis is time. 
Type "?dataPreview" in your R console to see the help file.

```{r, eval=TRUE, warning=FALSE}
# Create a vector for the time points.
tps = c(2000,2001,2002,2003,2005)

# I know the units of my varibel is number of pixels, thus I will create a sting for the vertical units.
vert_units <- "number of pixels"

# Next, pass the variables to the dataPreview function.
# NB: xAngle sets the orientation of the horizontal axis labels. See help file for more details.
dataPreview(rasstackY,
            timepoints = tps,
            vertunits = vert_units,
            xAngle = 0)
```


## Calling the `presenceData` function
The **presenceData** function creates data for the variable's number of presence and change. Type "?presenceData" in your R console to see the help file.
Output from the **presenceData** function serves as input data for the **presencePlot** function.
```{r, eval=TRUE, warning=FALSE}
# The function takes tw parameters: a raster stack and the value to flag as NA the default is -999
# Let's create the data for the rasstackY data
num_pres_change <- presenceData(rasstackY, nodata = -999)

# Let's take a look.
num_pres_change
```

The **presencePlot** function creates a map of the number of presence and changes for the variable of interest. 
Type "?presencePlot" in your R console to see the help file.
Bilintoh, Thomas Mumuni, and Robert G Jr. Pontius. “Methods to Compare Sites Concerning a Category's Change During Various Time Intervals.” forthcoming.
provide more information about the eight types of trajectories and the concept of the unified size
```{r, eval=TRUE, warning=FALSE}
# Let's pass thenum_pres_change we created to the presencePlot function.
# 4326
presencePlot(num_pres_change,
             pltunit = "m",
             dataEpsg = 32632,
             scalePos = "bottomright",
             narrowPos = "topleft",
             narrowSize = 1,
             categoryName = "forest",
             xAxis = "Longitude (m)",
             yAxis = "Latitude (m)",
             axisText = 1,
             axisLabel = 1,
             plotTitle = 1.2)

```

The **rastertrajData** function creates the data which serves as input for the **trajPlot** function. The output data comprises 
a raster map, pie chart, and corresponding attribute information such as colors and size.
Type "?rastertrajData" in your R console to see the help file.

```{r, eval=TRUE, warning=FALSE}
# Let's cteae trajectory data foe the rasstackY data.
# zeroabsence is a string of "yes" or "no" indicating if 0 means absence. The default is "yes".
# Type "?presencePlot"  in your R console to see the help file that describes all the parameters in the function.
traj_data <- rastertrajData(rasstackY,
                            zeroabsence = 'yes')
# Let's take a look.
traj_data
```

The **trajPlot** function creates the data which serves as input for the **trajPlot** function. The output data comprises 
a raster map, pie chart, and corresponding attribute information such as colors and size.
Type "?rastertrajData" in your R console to see the help file.

```{r, eval=TRUE, warning=FALSE}
# pass traj_data we created to the trajPlot function.
# Apart from the input, all the parameters are the same as the parameters for the presencePlot function
trajPlot(traj_data,
         axisShow = "no",
         categoryName = "forest",
         narrowPos = NA,
         scalePos = NA,
         scaleSize = 1.5,
         axisText = 1.2,
         axisLabel = 1.2,
         plotTitle = 1,
         legendTex = 1,
         xAxis = "Longitude (m)",
         yAxis = "Latitude (m)",
         downsample = TRUE)
```

The **rasterstackData** function creates results that serve as input for the **stackbarPlot** function. The output data comprises 
data frames and corresponding attribute information such as colors and unified size, average gross gain, and average gross loss. 
Type "?rasterstackData" in your R console to see the help file.
```{r, eval=TRUE, warning=FALSE}
# Let's cteae stacked bar data foe the rasstackY data.
# Create a vector for the time points.
tps = c(2000,2001,2002,2003,2005)

# I know the units of my variable is number of pixels, thus I will create a sting for the vertical units.
stackbar_data <- rasterstackData(rasstackY,
                                 timePoints = c(2000,2001,2002,2003,2005),
                                 spatialextent = 'unified',
                                 zeroabsence = 'yes',
                                 annualchange = 'yes',
                                 categoryName = 'variable',
                                 regionName = 'region',
                                 varUnits = "(squre kilometers)",
                                 constant  = 1)
# Let's take a look.
stackbar_data
```
The **stackbarPlot** function creates stack bar plots showing gross losses and gains of the trajectories and the three change components.
Type "?stackbarPlot" in your R console to see the help file.

```{r, eval=TRUE, warning=FALSE}
# pass stackbar_data we created to the stackbarPlot function.

# axisSize is a numerical value that control the size of the labels on tick marks of the horizontal and vertical tick marks.
# lbAxSize is a numerical value to control the size of the labels on the horizontal and vertical axis.
# lgSize is a numerical value to control the size of the legend text.
# titleSize is a numerical value to control the size of the title text.
# datbreaks is a string of "yes" or "no", which controls the range and sub-division of the vertical axis
# of the stacked bar plots.The default is "no", which automatically generates the range and interval of the vertical axis.
# If "no" the user need to mannual input values for "upperlym","lowerlym", "lymby","upperlym2", and "lymby2".
# pperlym if datbreaks set to "yes," is a numerical value to control the upper limit of the trajectory stack bar plot.
# lowerlym if datbreaks set to "yes," is a numerical value to control the lower limit of the trajectory stack bar plot.
# lymby if datbreaks set to "yes," is a numerical value to control interval on the vertical axis of the components of change stack bar plot.
# upperlym2 if datbreaks set to "yes," is a numerical value to control the upper limit of the components of change stacked bar plot.
# lymby2 if datbreaks set to "yes," is a numerical value to control the interval on the vertical axis of the components of change stacked bar plot.
# xAngle  is a numerical value to control the orientation of the text on the vertical axis of the trajectory stack bar plot.
stackbarPlot(stackbar_data,
             axisSize = 10,
             lbAxSize = 10,
             lgSize = 7.5,
             titleSize = 12,
             datbreaks = "no",
             upperlym = 35,
             lowerlym = - 50,
             lymby = 5,
             upperlym2 = 0.5,
             lymby2 = 0.1,
             xAngle = 0)
```

## References
Pontius Jr, R. G. (2022). Metrics That Make a Difference: How to Analyze Change and Error. Springer Nature Switzerland AG.

Bilintoh, T.M., Korah, A., Opuni, A., & Akansobe, A., (2022). Comparing the Trajectory of Urban Impervious Surface in Two Cities: 
The Case of Accra and Kumasi , Ghana. Land,  12 (927), 14. 

Pontius Jr, R. G., Krithivasan, R., Sauls, L., Yan, Y., & Zhang, Y. (2017). Methods to summarize change among land
categories across time intervals. Journal of Land Use Science, 12(4), 218–230. 


